Metric Spaces with Expensive Distances

### 2019-12-04 paper/draft_socg.tex
Typo
add more data

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of images
equipped with the Wasserstein (or the Earth Mover's) metric \cite{rubner2000earth}.

@@

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
a {$(1+\eps)$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated the \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
Our experiments with the Earth Mover's distance and persistence
diagrams of 3-D shapes show that blind greedy spanner can speed up computations
by a factor of 2 to 5, saving hundreds of hours of machine time,
even though the inputs in our data sets are of moderate size.
%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
%computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

\section{Background}
%
\myparagraph{Doubling dimension}
A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
if every ball of radius $r$ can be covered by at most $k$ balls of radius $r/2$,
and $k$ is the smallest number having that property.
The \textit{doubling dimension} of a doubling space is defined as $\log k$
(%since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
we always use $\log$ to denote the logarithm with base 2).
A subspace of a space with doubling dimension $d$ 
is always of doubling dimension $\leq 2d$.

@@

@@

On the other hand, spanner constructions based on Well-Separated Pair Decomposition\footnote{ 
for convenience, we repeat the definition of WSPD in \cref{sec:appendix_background}}~\cite{cal-kos-wspd,hm-fast} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a \espanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

\section{Blind spanners}
%
We introduce a new framework for constructing spanners
which we call \emph{blind spanners}: the idea is to maintain,
for every pair of points $(p_i,p_j)$,
a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
initially set to $0$ and $\infty$, respectively. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
we pick one of them, compute its distance and update the lower and upper bounds of
all pairs with respect to the newly acquired information. Here is the pseudocode:

@@

We performed three types of experiments: on points sampled from a low-dimensional Euclidean
space, on images with Earth Mover distance, and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

\begin{table}\centering
    \ra{1.2}
\begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
%\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
%\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
%\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
%\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
\bottomrule
\end{tabular}
    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
 with Wasserstein distance $W_q$.}
\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{c} & \multicolumn{2}{c}{$q=2$} & \phantom{c} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
%$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
%$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
%$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
%$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
%$\eps = 0.01$    &   4093  &  0  &&  5331  &  0  &&  16852  &  0\\
$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  11781  &  13557\\
$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  10541   &  9383\\
$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  9418    &  5146\\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
        \cmidrule{2-3} \cmidrule{5-7} 
                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
$\eps = 0.1$    &   $0.19$  &  $0.42$    &&  $618,587$   &  $262,770$    & $0.42$   \\
$\eps = 0.2$    &   $0.11$  &  $0.30$    &&  $618,587$   & $196,309$    & $0.31$   \\
$\eps = 0.5$    &   $0.04$  &  $0.18$    &&  $618,587$   & $112,995$    & $0.18$   \\
\bottomrule
\end{tabular}
    \caption{Results of \bgrdy algorithm on the dataset of 250 images
    with Earth Mover Distance, time in seconds.}
\label{tbl:emd_gtsrb}
\end{table}

@@

For experiments with images, we used the German traffic sign
dataset \cite{gtsrb} and the code for computing the Earth Mover distance
by Doran \cite{doran_pyemd}.
We randomly selected 250 images from the benchmark.
The first two columns of \cref{tbl:emd_gtsrb}
compare the sparseness of \bgrdy and greedy spanners.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
For $10 \%$ approximation, we can reduce
the running time by more than a factor of $2$,
saving $98$ hours of computation time.
For $20 \%$ approximation,
we speed up the computations by a factor of 3.

@@

\section{Additional Definitions}
\myparagraph{Well-Separated Pair Decomposition}
\label{sec:appendix_background}
Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

@@

\myparagraph{Earth Mover distance}
Let $\mu_P$ and $\mu_Q$
be two probability measures supported
on finite sets $P = \{p_1, \dots, p_n \}$
and $Q = \{ q_1, \dots, q_m \}$.
An $m \times n$ matrix $C$
is called a \textit{transportation plan},
if it satisfies the following three conditions:
\begin{enumerate}
    \item $c_{i,j} \geq 0$ for all entries $c_{i, j}$.
    \item $\sum_{i = 1}^m c_{i, j} = p_j$ for all $j = 1, \dots, n$.
    \item $\sum_{j = 1}^n c_{i, j} = q_i$ for all $i = 1, \dots, m$.
\end{enumerate}
%By definition, the \textit{cost} of transportation plan $C$
Let $\dist(\cdot, \cdot)$ denote the Euclidean
distance in plane.
The earth mover distance between $\mu_P$ and $\mu_Q$
is defined as
\[
    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j),
\]
where $\inf$ is taken over all transportation
plans $C$. Intuitively, $c_{i,j}$ is the
mass concentrated at the point $p_i$ that
is sent to the point $q_j$, and the earth mover
distance is the smallest total amount of work
needed to transform the measure $\mu_P$
into the measure $\mu_Q$.

Let $I_1$, $I_2$ be two greyvalue images.
We divide each greyvalue of $I_{1,2}$
by the sum of all greyvalues in the corresponding image.
Thus we can interpret $I_i$
as a probability measure supported on a rectangle
that has one vertex at the origin and
fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
The earth mover distance between images $I_1$ and $I_2$
is, by definition, the earth mover distance between
these probability measures.

>>>>

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even a constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of images
equipped with the Wasserstein (or the Earth Mover's) metric \cite{rubner2000earth}.

@@

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct an $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
a {$(1+\eps)$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated the \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance between a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.

Our experiments with the Earth Mover's distance and persistence
diagrams of 3-D shapes show that blind greedy spanner can speed up computations
by a factor of 2 to 5, saving hundreds of hours of machine time,
even though the inputs in our data sets are of moderate size.
%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
%computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithm gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

\myparagraph{Outline}
We review the properties of spanner algorithms in \cref{sec:back}.
In \cref{sec:blind_spanners}, we introduce different
variants of the blind spanner algorithm. Experimental
results on spanners are presented in \cref{sec:experiment_spanners}.
Then we move to the ANN problem, giving
an outline of the proof of the $O(\log n)$ bound in \cref{sec:ann}.
The full proof can be found in the appendix.
We report our experimental results for ANN in \cref{sec:experiment_ann}.

\section{Background}
\label{sec:back}
%
\myparagraph{Doubling dimension}
A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
if every ball of radius $r$ can be covered by at most $k$ balls of radius $r/2$,
and $k$ is the smallest number having that property.
The \textit{doubling dimension} of a doubling space is defined as $\log k$
(%since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
we always use $\log$ to denote the logarithm with base 2).
A subspace of a space with doubling dimension $d$ 
is always of doubling dimension $\leq 2d$.

@@

\myparagraph{Well-Separated Pair Decomposition}
Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

@@

On the other hand, spanner constructions based on Well-Separated Pair Decomposition~\cite{cal-kos-wspd,hm-fast} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a \espanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

\section{Blind spanners}
\label{sec:blind_spanners}
%
We introduce a new framework for constructing spanners
which we call \emph{blind spanners}: the idea is to maintain,
for every pair of points $(p_i,p_j)$,
a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
initially set to $0$ and $\infty$, respectively. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
we pick one of them, compute its distance and update the lower and upper bounds of
all pairs with respect to the newly acquired information. Here is the pseudocode:

@@

We performed three types of experiments: on points sampled from a low-dimensional Euclidean
space, on images with Earth Mover's distance, and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

\begin{table}\centering
    \ra{1.2}
\begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
$\eps = 0.4$               & 0.03 &  0.13          && 0.05 & 0.26        &&   0.06  & 0.34 \\
$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
$\eps = 0.8$               & 0.01 & 0.08           && 0.02  & 0.13        &&   0.02  & 0.17 \\
$\eps = 1.0$               & 0.01 &  0.06  && 0.01 & 0.10        &&   0.02  & 0.14 \\
\bottomrule
\end{tabular}
    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
 with Wasserstein distance $W_q$.}
\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{c} & \multicolumn{2}{c}{$q=2$} & \phantom{c} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
$\eps = 0.1$    &   $2605$  &  $1266$  &&  $3719$  &  $4085$  &&  $11$\,$781$  &  $13$\,$557$\\
$\eps = 0.2$    &   $2209$  &  $878$  &&  $3338$  &  $2574$  &&  $10$\,$541$  &  $9383$ \\
$\eps = 0.4$    &   $1988$  &  $586$  &&  $3088$  &  $1590$  &&  $-$    &  $6383$ \\
$\eps = 0.5$    &   $1744$  &  $483$  &&  $2835$  &  $1233$  &&  $9418$    &  $5146$ \\
$\eps = 0.8$    &   $1691$  &  $368$  &&  $2809$  &  $814$  &&  $-$    &  $3419$ \\
$\eps = 1.0$    &   $1645$  &  $324$  &&  $2780$  &  $672$  &&  $-$    &  $-$ \\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
        \cmidrule{2-3} \cmidrule{5-7} 
                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
$\eps = 0.1$    &   $0.20$  &  $0.43$    &&  $853\,682$   &  $248\,192$    & $0.29$   \\
$\eps = 0.2$    &   $0.11$  &  $0.30$    &&  $853\,682$   & $192\,302$    & $0.22$   \\
$\eps = 0.3$    &   $0.07$  &  $0.24$    &&  $853\,682$   & $161\,007$    & $0.19$   \\
$\eps = 0.4$    &   $0.05$  &  $0.20$    &&  $853\,682$   & $140\,470$    & $0.16$   \\
$\eps = 0.5$    &   $0.04$  &  $0.17$    &&  $853\,682$   & $129\,489$    & $0.15$   \\
$\eps = 0.8$    &   $0.02$  &  $0.13$    &&  $853\,682$   & $104\,851$    & $0.12$   \\
$\eps = 1.0$    &   $0.02$  &  $0.11$    &&  $853\,682$   & $92\,462$    & $0.11$   \\
\bottomrule
\end{tabular}
    \caption{Results of \bgrdy algorithm on the dataset of 250 images
    with Earth Mover's Distance, time in seconds.}
\label{tbl:emd_gtsrb}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
        \cmidrule{2-3} \cmidrule{5-7} 
                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
$\eps = 0.2$    &   $0.34$  &  $0.81$    &&  $185\,863$   & $133\,194$    & $0.72$   \\
$\eps = 0.3$    &   $0.20$  &  $0.69$    &&  $185\,863$   & $105\,723$    & $0.56$   \\
$\eps = 0.4$    &   $0.13$  &  $0.58$    &&  $185\,863$   & $86\,283$    & $0.46$   \\
$\eps = 0.5$    &   $0.09$  &  $0.50$    &&  $185\,863$   & $72\,447$    & $0.38$   \\
$\eps = 0.8$    &   $0.05$  &  $0.35$    &&  $185\,863$   & $48\,870$    & $0.26$   \\
$\eps = 1.0$    &   $0.04$  &  $0.30$    &&  $185\,863$   & $40\,955$    & $0.22$   \\
\bottomrule
\end{tabular}
    \caption{Results of \bgrdy algorithm on the dataset of 200 images
    with Earth Mover's Distance on \textit{Lab} histograms, time in seconds.}
\label{tbl:emd_gtsrb}
\end{table}

@@

For experiments with images, we used the German traffic sign
dataset GTSRB \cite{gtsrb} and the code for computing the Earth Mover's distance
by Doran \cite{doran_pyemd}.
We randomly selected 250 images from the benchmark.
The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
The first two columns compare the sparseness of \bgrdy and greedy spanners.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
If a $1.5$ approximation is acceptable,
then the \bgrdy algorithm speeds up
the computation by a factor of more than $5$,
and even for a $10 \%$ approximation, it reduces
the running time by more than a factor of $2$,
saving $98$ hours of computation time.

@@

\section{Additional Definitions}
\label{sec:appendix_background}

@@

\myparagraph{Earth Mover's distance}
Let $\mu_P$ and $\mu_Q$
be two probability measures supported
on finite sets $P = \{p_1, \dots, p_n \}$
and $Q = \{ q_1, \dots, q_m \}$.
An $m \times n$ matrix $C$
is called a \textit{transportation plan},
if it satisfies the following three conditions:
\begin{enumerate}
    \item $c_{i,j} \geq 0$ for all entries $c_{i, j}$.
    \item $\sum_{i = 1}^m c_{i, j} = p_j$ for all $j = 1, \dots, n$.
    \item $\sum_{j = 1}^n c_{i, j} = q_i$ for all $i = 1, \dots, m$.
\end{enumerate}
%By definition, the \textit{cost} of transportation plan $C$
Let $\dist(\cdot, \cdot)$ denote the Euclidean
distance in plane.
The Earth Mover's distance between $\mu_P$ and $\mu_Q$
is defined as
\[
    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j),
\]
where $\inf$ is taken over all transportation
plans $C$. Intuitively, $c_{i,j}$ is the
mass concentrated at the point $p_i$ that
is sent to the point $q_j$, and the Earth Mover's
distance is the smallest total amount of work
needed to transform the measure $\mu_P$
into the measure $\mu_Q$.

Let $I_1$, $I_2$ be two greyvalue images.
We divide each greyvalue of $I_{1,2}$
by the sum of all greyvalues in the corresponding image.
Thus we can interpret $I_i$
as a probability measure supported on a rectangle
that has one vertex at the origin and
fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
The Earth Mover's distance between images $I_1$ and $I_2$
is, by definition, the Earth Mover's distance between
these probability measures.
